Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-10-07 manuscript/main.tex
updated ncp results

\section{Results}
    1000 and 5000 SBC iterations are run for centered and reparameterised versions of the stocashtic volatility model. Rank statistics histograms are presented for the static parameters and the 1st, 500th and 1000th latent state parameters with 20 bins constructed for each parameter. The black horizontal line represents uniformity conditional on the number of bins and SBC iterations. The $\chi^2$ statistic is used to summarise the deviations from uniformity for all parameters since it is unrealistic to visually inspect all the histograms. The SBC algorithm is summarised below:

@@

    The centered parameterisation in particular displays a lack of uniformity for $\phi$ and $\sigma^2$. Rank statistics for $\sigma^2$ show two peaks of either end of the histogram and a left peak for $phi$. Posterior estimates of $\sigma^2$ are \textit{underdispersed} relative to the true parameter $\sigma^2_{sim}$ generated by the prior distribution. On the other hand, the posterior samples of $\phi$ are \textit{over estimating} the true parameter generated by the prior. This suggests the posterior geometry for these parameters (at least) are difficult to sample for the HMC algorithm.

@@

    Estimates of $\sigma^2$ and $\phi$ are more uniform under the reparameterised model. Instead, we observe a peak on the left side for the 500th state parameter $h_{500}$. Overall the estimates for non centered parameterisation are more favourable.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_1k.png}
        \caption{1000 SBC iterations for non centered model using Hamiltonian Monte Carlo. The rank statistic distributions for $\sigma^2$ and $\phi$ have improved. There is some evidence of over estimation of the 500th state parameter.}
        \label{fig:ncphmc1k}
    \end{figure}

@@

    Centered parameteristion posess similar improvement for $\mu$ and the state parameters. However, there are still major peaks at both ends of $\sigma^2$ and $\phi$ suggesting underdispersion of the posterior estimates relative to the prior distribution.

    These results suggests that HMC overall struggles to sample from the centered parameteristion of the stochastic volatility model. The reparameterised model results in much better and consistent inference for this sampling strategy, conditional on the model.

@@

    Further to calibration, the HMC algorithm struggles to efficiently generate posterior samples from the centered parameterisation of stochastic volatility. Figure \ref{fig:hmcess} visualises the distribution of ESS for the static parameters after 5000 SBC iterations. The non centered model appears to generate independent samples much more efficiently than the centered model. This is consistent with HMC struggliing to sample overall from the posteriors of $\sigma^2$ and $\phi$.

@@

    \subsection{Gaussian Mixture Approximation}
    The Gaussian Mixture Approximation is run for 1000 and 5000 SBC iterations with the first 10,000 samples discarded as burnin (consistent with \citet{kim1998stochastic}). However, unlike KSC who take 750,000 posterior draws from the model, the number of posterior samples taken as part of the SBC procedure is 9,999. Taking 750,000 draws from the Gaussian mixture approximation would lead to issues given the disk space and compute time required from run SBC. Furthermore, the number of posterior samples is larger than the 999 taken for the SBC experiment run for Hamiltonian Monte Carlo. Earlier experiments attempted 999 post burn in draws for the Gaussian approximation, however there are concerns that such a short chain would not have converged onto the target posterior distribution. Therefore 9,999 draws was chosen as a compromise under both constraints.

    The centered Gaussian mixture approximation contains issues with calibration. Figure \ref{fig:cpksc1k} presents the results from SBC where there is a lack of uniformity across all parameters. In particular, there is a large peak on the left hand side for $\mu$ and the right hand hand side of $\phi$. This suggests over estimation and under estimation of the true parameter respectively. All other parameters do not have any key features worth noting beyond noisy estimates around the true discrete uniform value.

@@

    Increasing the SBC iterations to 5000 does not improve the results. $\sigma^2$ and the state parameters contain arguably less noisy estimates around the uniform distribution. However, the skewed shapes of $\mu$ and $\phi$ adds further evidence that the sampling strategy is not returning the correct posterior estimates. This can be seen in Figure \ref{fig:cpksc5k}.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_cp_5k.png}
        \caption{5000 SBC iterations for centered Gaussian mixture approximation model. Estimates for $\sigma^2$ and the state parameters have marginally improved around the uniformity. However, there is no improvement in $\mu$ or $\phi$, suggesting a lack of calibration from the MCMC algorithm.}
        \label{fig:cpksc5k}
    \end{figure}

    Furthermore, the ESS estimates for this model indicate a high degree of autocorrelation and difficulty in generating independent samples from the posterior distribution. The median ESS for $\mu$, $\phi$ and $\sigma^2$ are 559, 103, 39.5 respectively for 10,000 post burn in draws. Despite having 10 times the number of draws, the number of effectively independent samples is much smaller relative to HMC, suggesting that this bespoke sampling strategy is highly inefficient.

@@

    The reparameterised model using the bespoke sampling strategy fails to generate sufficient results due to numerical errors and highly nonstationary chains. Figure X shows the traceplot from one simulation from the reparamterised model. The chain for $\phi$ is stuck to a value very close to 1 (unit root) and $\mu$ and $\sigma^2$ continuously increases in value until a numerical error is reached. Note that this occurs during the burnin phase of the markov chain. As a result, SBC is unable to be performed on this parameterisation of the model.

    There are a variety of reasons why this may happen. Perhaps the underlying software used to perform the sampling and kalman filter is not suitable for this paramterisation, or perhaps more broadly the MCMC strategy does not perform well with this parameterisation of the model. More work is required to diagnose the cause of this problem. Other avenues to improve the sampling of this model could be to rewrite the sampler from scratch or consider a different sampling method. Note that this is only one kind of reparamterisation and further research could explore other reparamterisations, such as non centered in scale.

    \subsection{Reweigthing rank statistics of Gaussian Mixture}
    \citet{kim1998stochastic} correct for approximation error in their method by using importance sampling weights. The reweighting procedure ensures that samples are drawn from the correct posterior density. This is applied to the calculation of the expected value of the posterior density to get a weighted mean estimate from the samples drawn from the Gaussian mixture approximation. The rank statistic can be written as a function of the expected value, so the reweighting step can be applied to the calculation of the rank statistics to see if this correction improves the calibration of the sampling approach.

    Let $v(\theta, h)$ be the log weights defined as the log difference between the posterior densities of the true model with log chi squared errors $log\: g(\theta, h | y^{\ast}_t)$ and the Gaussian mixture model $log\:  k(\theta, h_t | y^{\ast}_t)$.

@@

    This gives the importance sampling weight. The expectation for any function of the posterior samples can be written as a function of the weights.

@@

    The results from applying this reweighting step to the rank statistics are given on Figure \ref{fig:reweight1k}. There are no major improvements to the uniformity of rank statistics. Increasing the SBC iterations to 5000 as seen in Figure \ref{fig:reweight5k} also results in no major improvements. The shape of both sets of histograms is consistent with the shape of the unweighted rank statistics.

>>>>

\section{Results}
    1000 and 5000 SBC iterations are run for centered and reparameterised versions of the stochastic volatility model. Rank statistic histograms are constructed with 20 bins for the static parameters and the 1st, 500th and 1000th latent state variables. The black horizontal line represents uniformity conditional on the number of bins and SBC iterations. The $\chi^2$ statistic is used to summarise the deviations from uniformity for all parameters since it is unrealistic to visually inspect all the histograms. The SBC algorithm is summarised below:

@@

    The centered parameterisation in particular displays a lack of uniformity for $\phi$ and $\sigma^2$. Rank statistics for $\sigma^2$ and $\phi$ show two peaks of either end of the histogram. The posterior estimates of $\sigma^2$ and $\phi$ are \textit{underdispersed} relative to the prior distribution. This implies the posterior geometry of these parameters may be difficult to sample for the HMC algorithm.

@@

    Estimates of $\sigma^2$ and $\phi$ are more uniform under the reparameterised model. Instead, we observe a peak on the left side for the 500th state variable $h_{500}$. Overall the estimates for non centered parameterisation are more favourable.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/hmc_ncp_1k.png}
        \caption{1000 SBC iterations for non centered model using Hamiltonian Monte Carlo. The rank statistic distributions for $\sigma^2$ and $\phi$ have improved. There is some evidence that the posterior of the 500th state variable is overestimating the distribution of its true value.}
        \label{fig:ncphmc1k}
    \end{figure}

@@

    Centered parameteristion posess similar improvement for $\mu$ and the state variables. However, there are still major peaks at both ends of $\sigma^2$ and $\phi$ suggesting underdispersion of the posterior estimates relative to the prior distribution.

    These results suggests that HMC struggles to sample some parameters from the centered parameteristion of the stochastic volatility model. The reparameterised model possess stronger evidence of calibration for the parameters examined, implying more reliable and consistent inference.

@@

    Additionally, HMC struggles to generate efficient posterior samples from the centered parameterisation of the stochastic volatility model. Figure \ref{fig:hmcess} visualises the distribution of ESS for the static parameters after 5000 SBC iterations. The non centered model appears to generate independent samples much more efficiently than the centered model. This is consistent with HMC struggliing to sample overall from the posteriors of $\sigma^2$ and $\phi$.

@@

    \subsection{Gaussian Mixture Approximation}
    The Gaussian Mixture Approximation is run for 1000 and 5000 SBC iterations with the first 10,000 samples discarded as burnin (following the method of \citet{kim1998stochastic}). However, unlike KSC who take 750,000 posterior draws from the model, the number of posterior samples taken as part of the SBC procedure is 9,999. Taking 750,000 draws from the Gaussian mixture approximation would lead to issues due to disk space limitations and computational resources required to run SBC. Furthermore, the number of posterior samples is larger than the 999 taken for the SBC experiment run for Hamiltonian Monte Carlo. Earlier experiments attempted 999 post burn in draws for the Gaussian approximation, however there are concerns that such a short chain would not have converged onto the target posterior distribution. Therefore 9,999 draws was chosen as a compromise under both constraints.

    The both the centered and reparameterised Gaussian mixture model contains issues with calibration. Figure \ref{fig:cpksc1k} presents the results from the centered model. There is a lack of uniformity across all parameters. In particular, there is a large peak on the left hand side for $\mu$ and the right hand hand side of $\phi$. This suggests over estimation and under estimation of the prior distribution respectively. All other parameters do not have any key features worth noting beyond noisy estimates around the discrete uniform distribution.

@@

    The non centered model in location exhibits larger issues in calibration as shown in Figure \ref{fig:ncpksc1k}. Each histogram exhibits major peaks at one or both ends of the histgoram suggesting major issues in the sampler returning correct posterior estimates. The lack of uniformity for this model is distinctly due to a different shape. Whereas the lack of uniformity for previous SBC results with 1000 iterations may be due to noisy estimates.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{results/ksc_ncp_1k.png}
        \caption{1000 SBC iterations for non centered in location Gaussian mixture approximation model. The rank statistics for all selected parameters are non uniform in shape. The bespoke MCMC struggles to return the correct posteriors for this parametersation of the model.}
        \label{fig:ncpksc1k}
    \end{figure}

    Increasing the SBC iterations to 5000 does not improve the results for either parametersation. $\sigma^2$ and the state parameters contain arguably less noisy estimates around the uniform distribution in the centered parametersation. However, the skewed shapes of $\mu$ and $\phi$ adds further evidence that the sampling strategy is not returning the correct posterior estimates. This can be seen in Figure \ref{fig:cpksc5k}.

    Figure \ref{fig:ncpksc5k} shows no improvement to calibration results. The issues in calibration observed in Figure \ref{fig:ncpksc1k} are not due to noisy estimates and are likely the result of problems in the bespoke MCMC algorithm and model specification. Applying this parametersation and MCMC on real data will on balance produce incorrect posterior estimates.

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/ksc_cp_5k.png}
        \caption{5000 SBC iterations for centered Gaussian mixture approximation model. Estimates for $\sigma^2$ and the state parameters have marginally improved around the uniformity. However, there is no improvement in $\mu$ or $\phi$.}
        \label{fig:cpksc5k}
    \end{figure}

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.09]{results/ksc_ncp_5k.png}
        \caption{5000 SBC iterations for non centered Gaussian mixture approximation model. No improvement is observed in the distribution of rank statistics. The MCMC returns uncalibrated posterior estimates for this parametersation of the model.}
        \label{fig:ncpksc5k}
    \end{figure}

    The ESS estimates for this model indicate a high degree of autocorrelation and difficulty in generating independent samples from both parameterisations. The numerical results are summarised in table x. The estimate for $\mu$ in the reparameterised model exhibits a long right tail (longer than the 10,000 posterior draws suggesting the presence of negatively autocorrelated draws). Overall, the bespoke MCMC is quite insufficient at sampling the static parameters.

    % The median ESS for $\mu$, $\phi$ and $\sigma^2$ are 559, 103, 39.5 respectively for 10,000 post burn in draws. Despite having 10 times the number of draws, the number of effectively independent samples is much smaller relative to HMC, suggesting that this bespoke sampling strategy is highly inefficient.

    % - reparam has massive long tail for mu

@@

    Overall the results from the KSC algorithm indicates some issues with calibration. The centered parameterastion appears to be more favourable for this sampling approach. Other avenues to improve the sampling of this model could be to consider a different bespoke sampling method or another parameterisation such as non centered in scale.

    There are a few potential reasons for the poor calibration results. It may be due to the inefficiency of the MCMC to generate the correct posteriors. Additionally, it could be due to the approximation of the actual stochastic volatility model not producing accurate parameter estimates. This is addressed using importance sampling weights by KSC and explored in the next section.

    \subsection{Correcting approximation error using importance weights}
    \citet{kim1998stochastic} correct for approximation error in their method by using importance weights. The reweighting procedure ensures that samples are drawn from the correct posterior density. This is applied to the calculation of the expected value of the stochastic volatility posterior density using samples drawn from the Gaussian mixture approximation.

    Importance weights are defined as the ratio of the joint posterior from the stochastic volatility model and the posterior distribution of the approximation. A weight is produced for each posterior sample generated by MCMC. These weights correct the samples from the approximate distribution by increasing or decreasing the contribution of that sample in the calculation of the expectation.

    These weights can be applied in an additional sampling step using Metropolis Hastings or Importance Resampling (also known as sampling-importance resampling or SIR) to produce samples from the target posterior. However, the rank statistic can also be written as a function of the weighted expectation of the indicator random variable. This reweighting step is applied to the calculation of the rank statistics to see if the correction improves the calibration of the MCMC sampler.

    \subsubsection{Reweigthing rank statistics of the Gaussian Mixture Model}
    Let $v(\theta, h)$ be the log weights defined as the log difference between the posterior densities of the true model with log chi squared errors $log\: g(\theta, h | y^{\ast}_t)$ and the Gaussian mixture model $log\:  k(\theta, h_t | y^{\ast}_t)$.

@@

    This gives the normalised importance weight. The expectation for any function of the posterior samples can be written as a function of these weights. Let $S(\theta)$ be an indicator random variable that is a function of the posterior samples, the expectation can be written as:

@@

    The results from applying this reweighting step to the rank statistics of the centered model are given on Figure \ref{fig:reweight1k}. There are no major improvements to the uniformity of rank statistics. Increasing the SBC iterations to 5000 as seen in Figure \ref{fig:reweight5k} also results in no major improvements. The shape of both sets of histograms is consistent with the shape of the unweighted rank statistics. The reweighted rank statistics for the reparameterised model exhibit the same behaviour and can be found in Appendix X.
